---
title: "Human and Machine Learning — Assignment 1: Gaussians, Categories, and Clusters"
subtitle: "R (no-GenJAX) stencil — SOLUTION KEY"
author: "Joseph Austerweil"
date: "Spring 2026 (due Fri Jun 5, 2026 at 8:00pm)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Instructor answer key — do not distribute to students.** Worked solution for
the R stencil `clusters_nosoln.Rmd`.

**Corresponding textbook chapters:** Tutorial 1 Ch 5 (Bayesian inference) and
[Tutorial 3 Ch 5 — Mixture Models](https://josephausterweil.github.io/probintro/intro2/05_mixture_models/).

```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Problem 1: Gaussian-Gaussian Conjugate Model

In class, we derived the posterior and predictive distributions for a
Gaussian-Gaussian model: a Gaussian likelihood with unknown mean and known
variance, and a Gaussian prior on the mean. The conjugate posterior and
posterior-predictive are:

\begin{align*}
{\rm Posterior:} & &  \mu & | x_1,\dotsc, x_N \sim {\rm N} \left( \frac{\mu_0 \sigma_0^{-2} + \sigma_x^{-2} \sum_{n=1}^N{x_n}   } {\sigma_0^{-2} + N \sigma_x^{-2} }, \left[ \sigma_0^{-2} + N \sigma_x^{-2} \right]^{-1}  \right)  \\
{\rm Prediction} & & x_{N+1} & | x_1, \dotsc, x_N \sim {\rm N} \left( \frac{ \mu_0 \sigma_0^{-2} + \sigma_x^{-2} \sum_{n=1}^N{x_n}  }{\sigma_0^{-2} + N \sigma_x^{-2} },  \left[ \sigma_0^{-2} + N \sigma_x^{-2} \right]^{-1} + \sigma_x^2 \right)
\end{align*}

For this problem, use $\mu_0=0$ and $\sigma_0^2 = 1$.

### 1. (a) Prior

```{r prior, echo=TRUE}
# SOLUTION
mu_0 <- 0
sigma_0_squared <- 1
mu_range <- seq(-4, 4, length.out = 1000)

prior_df <- data.frame(
  mu = mu_range,
  density = dnorm(mu_range, mu_0, sqrt(sigma_0_squared))
)

ggplot(prior_df, aes(x = mu, y = density)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_area(fill = "steelblue", alpha = 0.15) +
  labs(x = expression(mu), y = expression(p(mu)),
       title = expression(paste("Prior: N(", mu[0], ", ", sigma[0]^2, ")"))) +
  theme_minimal()
```

The prior is a standard normal centered at 0 with standard deviation 1. It
captures our baseline belief about $\mu$ before seeing any data.

### 1. (b) One Datum update

```{r conjugate-update, echo=TRUE}
# SOLUTION
# Conjugate Gaussian-Gaussian update.
# Returns a list with components post_mean, post_var, pred_var.
conjugate_update <- function(mu_0, sigma_0_squared, sigma_x_squared, data) {
  N <- length(data)
  sum_x <- sum(data)

  posterior_precision <- 1 / sigma_0_squared + N / sigma_x_squared
  post_mean <- (mu_0 / sigma_0_squared + sum_x / sigma_x_squared) / posterior_precision
  post_var  <- 1 / posterior_precision
  pred_var  <- post_var + sigma_x_squared

  list(post_mean = post_mean, post_var = post_var, pred_var = pred_var)
}
```

```{r one-datum, echo=TRUE}
# SOLUTION
mu_0 <- 0
sigma_0_squared <- 1
x_1 <- c(2)
sigma_x_squared_values <- c(0.25, 4)
plot_range <- seq(-4, 6, length.out = 1000)

plot_df <- bind_rows(lapply(sigma_x_squared_values, function(sx2) {
  fit <- conjugate_update(mu_0, sigma_0_squared, sx2, x_1)
  data.frame(
    value = rep(plot_range, 2),
    density = c(dnorm(plot_range, fit$post_mean, sqrt(fit$post_var)),
                dnorm(plot_range, fit$post_mean, sqrt(fit$pred_var))),
    type = rep(c("posterior", "predictive"), each = length(plot_range)),
    sigma_x_squared = sx2
  )
}))

ggplot(plot_df, aes(x = value, y = density, color = type)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = x_1, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ sigma_x_squared,
             labeller = label_bquote(sigma[x]^2 == .(sigma_x_squared))) +
  labs(x = "value", y = "density") +
  theme_minimal()
```

The posterior and predictive share the same mean but the predictive is wider by
exactly $\sigma_x^2$ — the irreducible per-observation noise. Increasing
$\sigma_x^2$ from 0.25 to 4 makes the posterior move *less* toward $x_1 = 2$ (a
diffuse likelihood lets the prior dominate), while making the predictive *much*
wider. The likelihood variance therefore has opposite-direction effects on the
two distributions.

### 1. (c) Multiple data update

```{r five-data, echo=TRUE}
# SOLUTION
mu_0 <- 0
sigma_0_squared <- 1
data <- c(2.1, 2.5, 1.4, 2.2, 1.8)
sigma_x_squared_values <- c(0.25, 4)
plot_range <- seq(-4, 6, length.out = 1000)

plot_df <- bind_rows(lapply(sigma_x_squared_values, function(sx2) {
  fit <- conjugate_update(mu_0, sigma_0_squared, sx2, data)
  data.frame(
    value = rep(plot_range, 2),
    density = c(dnorm(plot_range, fit$post_mean, sqrt(fit$post_var)),
                dnorm(plot_range, fit$post_mean, sqrt(fit$pred_var))),
    type = rep(c("posterior", "predictive"), each = length(plot_range)),
    sigma_x_squared = sx2
  )
}))

ggplot(plot_df, aes(x = value, y = density, color = type)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = mean(data), linetype = "dashed", alpha = 0.5) +
  facet_wrap(~ sigma_x_squared,
             labeller = label_bquote(sigma[x]^2 == .(sigma_x_squared))) +
  labs(x = "value", y = "density") +
  theme_minimal()
```

The data mean is 2.0 here, same as the single datum in Part 1(b), so the
posterior *means* are similar across the two parts. What changes is the
*variance*: with $N = 5$ the posterior precision $1/\sigma_0^2 + N/\sigma_x^2$
has a five-times-larger data term, so the posterior is markedly tighter — most
dramatically for $\sigma_x^2 = 0.25$ (posterior variance $\approx 0.048$). The
predictive variance is still dominated by $\sigma_x^2$: it asymptotes to
$\sigma_x^2$ no matter how much data we collect.

# Problem 2: Gaussian Mixture / Categorization

Data are generated by first picking a category $c \in \{1, 2\}$ with
$P(c = 1) = \theta$, then generating the datum from that category's Gaussian.
For all of Problem 2, $\mu_1 = -1$ and $\mu_2 = 1$.

### 2. (a) Derivation: Categorization

By **Bayes' rule** with prior $P(c_1 = 1) = \theta$ and likelihood
$p(x_1 \mid c_1 = 1) = {\rm N}(x_1; \mu_1, \sigma_1^2)$, the denominator given by
the **Law of Total Probability**:

\begin{equation*}
P(c_1 = 1 | x_1) = \frac{p(x_1 \mid c_1 = 1) P(c_1 = 1)}{p(x_1)}
= \frac{\theta {\rm N}(x_1; \mu_1, \sigma_1^2)}{\theta {\rm N}(x_1; \mu_1, \sigma_1^2) + (1-\theta){\rm N}(x_1; \mu_2, \sigma_2^2)}
\end{equation*}

### Answer:

Step 1 — Bayes' rule:
$P(c_1=1 \mid x_1) = \dfrac{p(x_1 \mid c_1=1)\,P(c_1=1)}{p(x_1)} = \dfrac{\theta\,{\rm N}(x_1;\mu_1,\sigma_1^2)}{p(x_1)}$.

Step 2 — expand the denominator with the Law of Total Probability:
$p(x_1) = \sum_c p(x_1 \mid c)P(c) = \theta\,{\rm N}(x_1;\mu_1,\sigma_1^2) + (1-\theta)\,{\rm N}(x_1;\mu_2,\sigma_2^2)$.

Step 3 — substitute back, giving the boxed result above.

### 2. (b) Categorization

```{r categorization, echo=TRUE}
# SOLUTION
mu_1 <- -1
mu_2 <- 1
theta_values <- c(0.5, 0.75)
configs <- list(c(1, 1), c(0.5, 2))
x_range <- seq(-6, 6, length.out = 1000)

rows <- list()
for (theta in theta_values) {
  for (cfg in configs) {
    sigma_1_sq <- cfg[1]; sigma_2_sq <- cfg[2]
    lik1 <- dnorm(x_range, mu_1, sqrt(sigma_1_sq))
    lik2 <- dnorm(x_range, mu_2, sqrt(sigma_2_sq))
    posterior_c1 <- (theta * lik1) / (theta * lik1 + (1 - theta) * lik2)
    rows[[length(rows) + 1]] <- data.frame(
      x = x_range, prob = posterior_c1,
      theta = theta,
      config = sprintf("sigma_1^2=%s, sigma_2^2=%s", sigma_1_sq, sigma_2_sq)
    )
  }
}
plot_df <- bind_rows(rows)

ggplot(plot_df, aes(x = x, y = prob,
                    color = factor(theta), linetype = config)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.5, alpha = 0.3) +
  labs(x = "x", y = expression(P(c[1] == 1 ~ "|" ~ x)),
       color = expression(theta), linetype = "configuration") +
  theme_minimal()
```

Changing $\theta$ and changing the variances do **not** have the same effect.
$\theta$ adds a constant offset to the log-odds — it shifts the whole curve
up/down and (with equal variances) moves the decision boundary, without
changing the curve's shape. The variances change the *shape*: when
$\sigma_1^2 \neq \sigma_2^2$ the log-odds is quadratic in $x$, so the curve is
no longer a simple monotone sigmoid.

### 2. (c) Derivation --- Prediction

By the **Law of Total Probability**:

\begin{equation*}
p(x_1) = \sum_c p(x_1 \mid c)P(c) = \theta {\rm N}(x_1; \mu_1, \sigma_1^2) + (1-\theta){\rm N}(x_1; \mu_2, \sigma_2^2)
\end{equation*}

### Answer:

$p(x_1) = \sum_c p(x_1 \mid c)\,P(c) = p(x_1 \mid c_1{=}1)P(c_1{=}1) + p(x_1 \mid c_1{=}2)P(c_1{=}2)$,
and substituting $P(c_1{=}1)=\theta$, $P(c_1{=}2)=1-\theta$ and the two Gaussian
likelihoods gives the boxed result.

### 2. (d) Prediction

```{r prediction, echo=TRUE}
# SOLUTION
mu_1 <- -1
mu_2 <- 1
theta_values <- c(0.5, 0.75)
configs <- list(c(1, 1), c(0.5, 2))
x_range <- seq(-6, 6, length.out = 1000)

rows <- list()
for (theta in theta_values) {
  for (cfg in configs) {
    sigma_1_sq <- cfg[1]; sigma_2_sq <- cfg[2]
    p_x <- theta * dnorm(x_range, mu_1, sqrt(sigma_1_sq)) +
           (1 - theta) * dnorm(x_range, mu_2, sqrt(sigma_2_sq))
    rows[[length(rows) + 1]] <- data.frame(
      x = x_range, density = p_x,
      theta = theta,
      config = sprintf("sigma_1^2=%s, sigma_2^2=%s", sigma_1_sq, sigma_2_sq)
    )
  }
}
plot_df <- bind_rows(rows)

ggplot(plot_df, aes(x = x, y = density,
                    color = factor(theta), linetype = config)) +
  geom_line(linewidth = 1) +
  labs(x = "x", y = "p(x)",
       color = expression(theta), linetype = "configuration") +
  theme_minimal()
```

$\theta$ controls the relative *heights* of the two component peaks; the
variances control their *widths*. With equal $\sigma^2$ the mixture is cleanly
bimodal; with $(\sigma_1^2, \sigma_2^2) = (0.5, 2)$ one component becomes a
narrow spike and the other a broad shoulder, and the visible bimodality can
wash out. They are not the same effect.

# Bonus (optional): Part 2(e) — verify the marginal by Monte Carlo

```{r bonus-monte-carlo, echo=TRUE}
# SOLUTION
set.seed(42)
N <- 2000
theta <- 0.7
mu_1 <- -1; mu_2 <- 1; sigma_1 <- 1; sigma_2 <- 1
x_range <- seq(-6, 6, length.out = 1000)

c_samples <- rbinom(N, 1, theta)            # 1 = category 1, 0 = category 2
x_samples <- ifelse(c_samples == 1,
                    rnorm(N, mu_1, sigma_1),
                    rnorm(N, mu_2, sigma_2))

analytical <- theta * dnorm(x_range, mu_1, sigma_1) +
              (1 - theta) * dnorm(x_range, mu_2, sigma_2)

ggplot(data.frame(x = x_samples), aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 60, alpha = 0.5,
                 fill = "steelblue") +
  geom_line(data = data.frame(x = x_range, density = analytical),
            aes(x = x, y = density), color = "red", linewidth = 1) +
  labs(title = "Monte Carlo p(x) vs. analytical mixture") +
  theme_minimal()
```

The simulated histogram matches the analytical $p(x)$ from Part 2(d) — an
empirical check of the Law of Total Probability.

## Submission

Submit this knitted .Rmd plus the derivations for Parts 2(a) and 2(c).
